Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-11 rebuttal.tex

The article over-simplifies concepts, such as molecular representations, for an audience which is perhaps even more well versed in graph theory than many cheminformaticians. The casual reader should be left with an understanding of specific challenges to which they can bring to bear methods from their field. Vague challenges are mentioned, but not in any clarity enough to go away and tackle them.

The manuscript provides a very specific overview on the aspects of the open source community in cheminformatics. Therefore, I would recommend that this be clarified in the title to avoid confusion or the manuscript broadened to
cover the actual extent of the field, which is far vaster than is covered here.

The authors make no (or very little) mention of the founding principles of cheminformatics. Very few references cover the algorithms that have been developed and published.

\comment{The manuscript is poor in relevant citations with none of the
  classical references provided, such as Morgan 1965, Weininger 1988,
  Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.}
\response{The text has been updated to include the ``classic''
  references. In addition the supplemental history document lists a
  number of milestones in cheminformatics and their relevant references.}

### 2011-11-11 rebuttal.tex

\comment{The authors state that 3D information is lost when only
  considering the molecular graph. However, they do not make it clear
  that the 3D space is implicitly encoded in the molecular graph –
  this is essential to cover properly.}
\response{We have update the text on page 2 to note that 3D
  information is implicit in 2D representations}

\comment{If this manuscript were to be published it would be
  completely rewriting the history of cheminformatics}
\response{While we agree with the reviewer that some key milestones in
  the history of cheminformatics were not included in the original
  article, we believe that the addition of the supplemental history
  addresses this problem. Furthermore, our original aim was \emph{not}
  to provide a complete historical overview of the field. Others have
  already provided such overviews. Our aim was to present a brief
  overview of \emph{topics} in cheminformatics, selected because we
  believed that they represent on going challenges for the field and
  thus amenable to coss-disciplinary efforts.}

\fbox{\textbf{Reviewer 3}}

\comment{The manuscript provides a very specific overview on the
  aspects of the open source community in cheminformatics. }
\response{As we have noted below, the current article is not a
  comprehensive historical review of the field. We have included a
  supplement  that does provide a very brief overview of historic
  milestones. Furthermore, we have updated the text to note commercial
  vendors and solutions; but we still stress on Open Source and Open
  Access solutions to allow reasders to explore the problems and
  topics we have discussed with a minimum of encumbrances}

### 2011-11-11 rebuttal.tex

\documentclass[11pt,a4paper]{article}
\usepackage{times}
\usepackage{fullpage}
\setlength{\parindent}{0cm}

\begin{document}
\newcommand{\comment}[1]{\textit{``\ldots #1''}\par\vspace{0.5em}}
\newcommand{\response}[1]{#1\vspace{1em}}

We would like to thank the anonymous reviewers for their constructive
comments. We have updated the main text in several places and have
also included a supplemental document that provides a brief
description of terminology and a list of some historic milestones in
the field. The following sections provide a detailed response to
individual reviewer comments.

### 2011-11-11 rebuttal.tex

\comment{Proteochemometric methods are mentioned a couple of times but
  these are largely un-noticed by the community currently. I do not
  think that it is necessarily appropriate to include these as
  tried-and-tested methods in an over-arching review such as this.}
\response{While it is true that proteochemometric methodologies are
  not as well known as say docking or pharmacophore modeling, we
  choose to include them since they are an example of a (statistical)
  QSAR technique that does explicitly include ligand and receptor features}

\comment{RDKit is not mentioned at all in this article and is widely
  used as a cheminformatics toolkit in industry and academia and free
  and open-source (although from commercial beginnings).}
\response{We agree that this toolkit should have been included. We
  have updated the text on page 7 to take note of RDKit, though we
  still primarily focus on the CDK and OpenBabel. However, Table 2
  includes RDKit in the feature comparison}

\comment{Figure 1: what is meant by 'closer to reality' This needs to
  be clarified. This figure and its caption requires a little more
  consideration to appropriately convey the message, which this
  reviewer takes to be that there are many useful ways to represent
  molecular structures.}
\response{We have updated the caption to remove the phrase ``closer to
reality'' since this is not the main point. Instead we have noted that
the representations contain differing amounts and types of information
that are equally valid, but suited for different purposes.}

### 2011-11-11 paper.tex
Added more responses

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce the
  new therapeutic molecules.  Only within the last decade---and as an
  effect of a change in mindsets, government-mandated policy changes
  and last but not least because of chemists volunteering their time
  for an Open Science ``movement''---have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding a cure for malaria and other ``neglected'' diseases.
\end{abstract}

@@

Making chemical information public is important because the drugs we
take are the result of intensive and costly research, which is
facilitated by the amount of information available. The more
information is shared (a process very hard to achieve in
pharmaceutical companies), the more information is available to each
researcher, facilitating the development of novel treatments. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available--- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
annotations such as synonyms, known targets, mode of action,
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

@@

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The top left hand corner
shows a structure a chemist would draw it---every ``corner''
representing a carbon atom (which can also be represented as a C),
every O denoting oxygen, every N nitrogen, every H hydrogen, and so
on. This is the standard format chemists use when exchanging
information with their peers.  However, this format is not directly
amenable to data mining---different atoms might be collapsed into one
node, and sets of atoms might have the same relevant properties
assigned. Hence, the top right hand corner of
Figure~\ref{figure:chemical-structures} shows an annotated graph
representation, taking into account that atoms, though being of
different elemental atom type, can have similar
properties---\emph{e.g.}, both oxygen and nitrogen carry an ``On''
label in this graph.

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{The many different facets of representing chemical structures. Shown
are a chemist's usual representation of a molecule (top left) and a
graph-based representation amenable to data mining (top right). These
representations are supplemented by a 3D representation of the same molecule
(bottom left) that is closer to reality and a representation using surface
properties (bottom right), which is often more relevant for the properties the
molecule exhibits in an experimental setup (such as when measuring its
solubility, or bioactivity).}
\label{figure:chemical-structures}
\end{figure}

@@

To extend the concept of three-dimensionality further, in many
biological systems the \emph{surface properties} of the molecule are
what appear to be relevant, as shown in the bottom right hand corner.
Hence, while graph mining on chemicals is certainly a promising avenue
to explore, one needs to keep in mind that not all molecular
properties are encoded in the graph (for example, neither rotational
states, nor different charge states are encoded) and thus often other
``descriptors'' need to be used to capture the essentials of a
chemical system \cite{bender2004}. Hence, one of the difficulties of a
cheminformatician is that the equivalence of particular atoms (or sets
of atoms) is usually not known from the onset but it rather depends on
the particular system considered---involving a large amount of
trial-and-error in the choice of a ``proper'' description for
molecules in a particular case.

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. If we oversimplify the nature of chemical structures and
their properties, we can not describe the world around us in
sufficient detail. On the other hand, if we try to encode every detail
(\emph{e.g.}, all possible 3D structures and their properties, all
metabolites of a structure in humans in different tissues), the
complexity of collecting, storing, and mining these data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists---where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data which achieve optimal balances between the different risks in
generalizable encodings and complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing enough open standards and
tools so software engineering concepts allow applications and data to
be used across heterogeneous platforms, and 3. mining the many
chemical property spaces in a time- and space-efficient way. We will
discuss all three goals in the following sections.

@@

There is considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. One problem is the annotation with the ``protein targets'' a molecule binds to---here dozens of
protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a cumbersome exercise.

@@

Enumerating molecules is not only an interesting academic exercise but
has practical utility as well. The foremost application of enumeration
is \emph{structure elucidation} of natural compounds such as
metabolites. Metabolites are small molecules that can be found an
organism (such as a human) and which in many cases indicate the state
the organism is in---such as indicating a disease, or a fasted state,
etc.---and with a sufficiently large database the idea is to be able
to detect the chemical composition of body fluids and to arrive at
conclusions regarding the particular state of the organism at this
point in time.  Ideally, the wishful bench chemist collects
experimental data from an unknown compound, the data is fed into a
code, and the resulting structure is returned. Although that
streamlined process may not yield a unique solution, there are
commercial software products such as MOLGEN (\url{http://molgen.de})
that can, for instance, list all structures matching a given molecular
formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to
design compounds (drugs, for instance) that optimize some physical,
chemical, or biological property or activity. Although less prolific
and more recent than structure elucidation, molecular design has
introduced some novel solutions to molecular enumeration. Finally,
with the advent of \emph{combinatorial chemistry} (the rapid synthesis
of different but related molecules), molecular enumeration has taken a
central role as it allows computational chemists to construct virtual
libraries and test hypotheses, and it provides guidance in the design
of optimal combinatorial experiments.

@@

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB}. Yet, the
current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates
molecules up to 13 heavy atoms---given the possibly exponential
increase in search space with the number of atoms, it seems unlikely
that the practically relevant chemical space of about 30 heavy atoms
will be exhaustively enumerated in the near future.

@@

It is important to note that certain challenges faced by nearly all
cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality---and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. It is important to realize that although
chemical representations have increased in sophistication over the
years, many chiral molecules cannot be satisfactorily handled by
current representation systems. Perhaps a more challenging aspect is
that the chemical representation of a molecule really needs to capture
non-static graphs, in order to take into account delocalisation and
the tautomeric phenomena molecules can undergo in practice.

@@

\caption{Some of the properties potential drug candidates need to have to be considered as ``promising,'' either in the different research and development stages, as well as a drug that is finally admitted to the market. Note that the particular requirements are also heavily project-dependent, and that also in cases where cheminformatics approaches can be employed in drug design an experimental confirmation of the prediction is required---the computer can hence decrease the experimental burden, but not replace it altogether.}

>>>>

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce the
  new therapeutic molecules.  Only within the last decade -- and as an
  effect of a change in mindsets, government-mandated policy changes
  and last but not least because of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Making chemical information public is important because the drugs we
take are the result of intensive and costly research, which is
facilitated by the amount of information available. The more
information is shared (a process very hard to achieve in
pharmaceutical companies), the more information is available to each
researcher, facilitating the development of novel treatments. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
annotations such as synonyms, known targets, mode of action,
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

@@

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The top left hand corner
shows a structure a chemist would draw it -- every ``corner''
representing a carbon atom (which can also be represented as a C),
every O denoting oxygen, every N nitrogen, every H hydrogen, and so
on. This is the standard format chemists use when exchanging
information with their peers.  However, this format is not directly
amenable to data mining -- different atoms might be collapsed into one
node, and sets of atoms might have the same relevant properties
assigned. Hence, the top right hand corner of
Figure~\ref{figure:chemical-structures} shows an annotated graph
representation, taking into account that atoms, though being of
different elemental atom type, can have similar
properties -- \emph{e.g.}, both oxygen and nitrogen carry an ``On''
label in this graph.

\begin{figure}
\centering
\includegraphics[height=2in]{chemstructs.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemists traditional 2D depiction (A), a labeled graph (B), a 3D
  conformation (C) and a surface represention (D). All show differing
  (and valid) aspects of the same molecule and are suited for
  differing purposes. The choice of representation is guided by the
  intended use of the structure data.}
\label{figure:chemical-structures}
\end{figure}

@@

To extend the concept of three-dimensionality further, in many
biological systems the \emph{surface properties} of the molecule are
what appear to be relevant, as shown in the bottom right hand corner.
Hence, while graph mining on chemicals is certainly a promising avenue
to explore, one needs to keep in mind that not all molecular
properties are encoded in the graph (for example, neither rotational
states, nor different charge states are encoded) and thus often other
``descriptors'' need to be used to capture the essentials of a
chemical system \cite{bender2004}. Hence, one of the difficulties of a
cheminformatician is that the equivalence of particular atoms (or sets
of atoms) is usually not known from the onset but it rather depends on
the particular system considered -- involving a large amount of
trial-and-error in the choice of a ``proper'' description for
molecules in a particular case.

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. If we oversimplify the nature of chemical structures and
their properties, we can not describe the world around us in
sufficient detail. On the other hand, if we try to encode every detail
(\emph{e.g.}, all possible 3D structures and their properties, all
metabolites of a structure in humans in different tissues), the
complexity of collecting, storing, and mining these data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists -- where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data which achieve optimal balances between the different risks in
generalizable encodings and complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing enough open standards and
tools so software engineering concepts allow applications and data to
be used across heterogeneous platforms, and 3. mining the many
chemical property spaces in a time- and space-efficient way. We will
discuss all three goals in the following sections.

@@

There is considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. One problem is the annotation with the ``protein targets'' a molecule binds to -- here dozens of
protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a cumbersome exercise.

@@

Enumerating molecules is not only an interesting academic exercise but
has practical utility as well. The foremost application of enumeration
is \emph{structure elucidation} of natural compounds such as
metabolites. Metabolites are small molecules that can be found an
organism (such as a human) and which in many cases indicate the state
the organism is in -- such as indicating a disease, or a fasted state,
etc. -- and with a sufficiently large database the idea is to be able
to detect the chemical composition of body fluids and to arrive at
conclusions regarding the particular state of the organism at this
point in time.  Ideally, the wishful bench chemist collects
experimental data from an unknown compound, the data is fed into a
code, and the resulting structure is returned. Although that
streamlined process may not yield a unique solution, there are
commercial software products such as MOLGEN (\url{http://molgen.de})
that can, for instance, list all structures matching a given molecular
formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to
design compounds (drugs, for instance) that optimize some physical,
chemical, or biological property or activity. Although less prolific
and more recent than structure elucidation, molecular design has
introduced some novel solutions to molecular enumeration. Finally,
with the advent of \emph{combinatorial chemistry} (the rapid synthesis
of different but related molecules), molecular enumeration has taken a
central role as it allows computational chemists to construct virtual
libraries and test hypotheses, and it provides guidance in the design
of optimal combinatorial experiments.

@@

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB}. Yet, the
current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates
molecules up to 13 heavy atoms -- given the possibly exponential
increase in search space with the number of atoms, it seems unlikely
that the practically relevant chemical space of about 30 heavy atoms
will be exhaustively enumerated in the near future.

@@

It is important to note that certain challenges faced by nearly all
cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality -- and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. It is important to realize that although
chemical representations have increased in sophistication over the
years, many chiral molecules cannot be satisfactorily handled by
current representation systems. Perhaps a more challenging aspect is
that the chemical representation of a molecule really needs to capture
non-static graphs, in order to take into account delocalisation and
the tautomeric phenomena molecules can undergo in practice.

@@

\caption{Some of the properties potential drug candidates need to have to be considered as ``promising,'' either in the different research and development stages, as well as a drug that is finally admitted to the market. Note that the particular requirements are also heavily project-dependent, and that also in cases where cheminformatics approaches can be employed in drug design an experimental confirmation of the prediction is required -- the computer can hence decrease the experimental burden, but not replace it altogether.}
